---
title: "Classification of Spotify Data"
subtitle: "DATA 606 - W2023 Final Project"
author: "Kane Smith, Rodrigo Rosales Alvarez, Arhur Trim, Jordan Keelan, Scott Bennett"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    extra_dependencies:
    - bbm
    - xcolor
---

\newpage
\tableofcontents
\pagebreak

```{r setup, include=FALSE}
# Set R Chunk options
knitr::opts_chunk$set(echo = TRUE)

# Set the number of significant digits
options(scipen=50, digits=4)

library(dplyr)
library(ggplot2)
libr

```


# 1 Introduction

## 1.1 Background

## 1.2 Objective & Topic Importance

# 2 Methodology

## 2.1 Software Used

## 2.2 Data Source

## 2.3 Summary of Variables

```{r}
artists_df <- data.frame(read.csv("../spotify_dataset/artists.csv"))
tracks_df <- data.frame(read.csv("../spotify_dataset/tracks.csv"))

```

```{r}
head(artists_df)
# head(tracks_df)

# glimpse(artists_df)

popular_df <- tracks_df %>%
                    filter(popularity > 1)

mean(tracks_df$popularity)
mean(popular_df$popularity)
                    

ggplot(data=tracks_df, aes(x=popularity)) + geom_histogram()
ggplot(data=artists_df, aes(x=popularity)) + geom_histogram()

tracks_df <- tracks_df %>%
              mutate(num_artists = length(as.list(strsplit(artists, ","))))

```

```{r}
x = "['Dick Haymes', 'Victor Young', 'His Orchestra']"

str_replace()
```



## 2.4 Data Cleaning

# 3 Genre Classification

## 3.1 LDA

### 3.1.1 Assumptions

## 3.2 Classification Tree

### 3.2.1 Assumptions

## 3.4 Summary of Findings

# 4 Classification - Hit or not?




## 4.1 Logistic Regression

### 4.1.1 Assumptions

## 4.2 Classification Tree

### 4.2.1 Assumptions

## 4.3 Summary of Findings

# 5 Conclusion

# 6 References

















