---
title: "Classification of Spotify Data"
subtitle: "DATA 606 - W2023 Final Project"
author: "Kane Smith, Rodrigo Rosales Alvarez, Arhur Trim, Jordan Keelan, Scott Bennett"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    extra_dependencies:
    - bbm
    - xcolor
---

\newpage
\tableofcontents
\pagebreak

```{r setup, include=FALSE}
# Set R Chunk options
knitr::opts_chunk$set(echo = TRUE)

# Set the number of significant digits
options(scipen=50, digits=4)

library(dplyr)
library(ggplot2)
library(tree)
library(MASS)
library(ISLR)
library(car)
library(sampling)
library(caret)
library(AppliedPredictiveModeling)
```


# 1 Introduction

## 1.1 Background

## 1.2 Objective & Topic Importance

# 2 Methodology

## 2.1 Software Used

## 2.2 Data Source

## 2.3 Summary of Variables

```{r}
artists_df <- data.frame(read.csv("../spotify_dataset/artists.csv"))
tracks_df <- data.frame(read.csv("../spotify_dataset/tracks.csv"))
```

```{r}
#head(artists_df)
# head(tracks_df)

# glimpse(artists_df)

popular_df <- tracks_df %>%
                    filter(popularity > 1)

#mean(tracks_df$popularity)
#mean(popular_df$popularity)
                    

#ggplot(data=tracks_df, aes(x=popularity)) + geom_histogram()
#ggplot(data=artists_df, aes(x=popularity)) + geom_histogram()

#tracks_df <- tracks_df %>%
#              mutate(num_artists = length(as.list(strsplit(artists, ","))))

popular_labels <- popular_df[, c("id", "name", "artists", "id_artists", "release_date")]
popular_factors <- popular_df[, c("explicit", "key", "mode", "time_signature")]
popular_numeric <- popular_df[, c("duration_ms","popularity", "danceability", "energy", "loudness", "speechiness", "acousticness", "instrumentalness", "liveness", "valence", "tempo")]
popular_scaled <- scale(popular_numeric)
popular_factors <- lapply(popular_factors,factor)
popular_df2 <- data.frame(popular_labels, popular_factors, popular_scaled)
colnames(popular_df2)
```

## 2.4 Data Cleaning

# 3 Genre Classification

## 3.1 LDA

### 3.1.1 Assumptions

## 3.2 Classification Tree
```{r}
set.seed(1)

N <- nrow(popular_df2)
n <- N * 0.8

idx = sample(1:N, size=n, replace = FALSE)
train = popular_df2[idx,]
test = popular_df2[-idx,]
```

```{r}
popularity_tree_fit <- tree(popularity ~ duration_ms+explicit+danceability+energy+key+loudness+mode+speechiness+acousticness+instrumentalness+liveness+valence+tempo+time_signature, data=train)
summary(popularity_tree_fit)
```
```{r}
popularity_tree_predict <- predict(popularity_tree_fit, test)

sqrt(mean((popularity_tree_predict - test$popularity)^2)) 
```
# Regression Tree
```{r}
set.seed(1)

N <- nrow(popular_df2)
n <- N * 0.8

idx = sample(1:N, size=n, replace = FALSE)
train = popular_df2[idx,]
test = popular_df2[-idx,]

popularity_tree_fit <- tree(popularity ~ duration_ms+explicit+danceability+energy+key+loudness+mode+speechiness+acousticness+instrumentalness+liveness+valence+tempo+time_signature, data=train)

summary(popularity_tree_fit)

popularity_tree_predict <- predict(popularity_tree_fit, test)

sqrt(mean((popularity_tree_predict - test$popularity)^2))
```

```{r}
plot(popularity_tree_fit)
text(popularity_tree_fit ,pretty =0)
```

```{r}
cv.popularity=cv.tree(popularity_tree_fit, K=10)
#plot(cv.popularity$popularity, cv.popularity$dev,type='b')
```

```{r}
plot(cv.popularity$size, cv.popularity$dev,type='b')
```


```{r}
ctrl <- trainControl(method = "cv", number = 10)

tree_model <- train(popularity ~ duration_ms+explicit+danceability+energy+key+loudness+mode+speechiness+acousticness+instrumentalness+liveness+valence+tempo+time_signature, data = popular_df, trControl = ctrl, method = "rpart")
print(tree_model)
```


### 3.2.1 Assumptions

## 3.4 Summary of Findings

# 4 Classification - Hit or not?
```{r}
popular_df2 <- popular_df2 %>% mutate(popularity_coded = ifelse(popularity >= quantile(popular_df2$popularity, c(.75))[1], 1, 0))


# Convert the outcome variable to a factor
popular_df2$popularity_coded <- as.factor(popular_df2$popularity_coded)

#popular_df$coded = rep("no hit", nrow(popular_df)) # default
#popular_df$coded[popular_df$popularity > 74] = "hit"
```

```{r}
quantile(popular_df2$popularity, c(.75))[1] + 1
```


```{r}
print(table(popular_df2$popularity_coded)) # 0=526964 , 1=2994
print(unique(popular_df2$popularity_coded))
```

```{r}
idx_lr = sampling:::strata(popular_df2, stratanames = c("popularity_coded"), size=c(400977/8,128981/8), method="srswor")

lr_train = popular_df2[idx_lr$ID_unit,]
lr_test = popular_df2[-idx_lr$ID_unit,]

lr_fit <- glm(popularity_coded ~ duration_ms+explicit+danceability+energy+key+loudness+mode+speechiness+acousticness+instrumentalness+liveness+valence+tempo+time_signature, data = lr_train, family=binomial)
summary(lr_fit)
```
```{r}
lr_results <- predict(lr_fit, lr_test, type="response")
m = table(lr_results, lr_test$popularity_coded) # comparing the actual vs predicted values
(m[1,2] + m[2,1]) / nrow(lr_test)
```
k_fold
```{r}
ctrl <- trainControl(method = "cv", number = 10)

#fit a regression model and use k-fold CV to evaluate performance
lr_kfold_model <- train(popularity_coded ~ duration_ms+explicit+danceability+energy+key+loudness+mode+speechiness+acousticness+instrumentalness+liveness+valence+tempo+time_signature, data = lr_train, method = "glm", trControl = ctrl, family="binomial")

print(lr_kfold_model)
```


## 4.1 Logistic Regression

### 4.1.1 Assumptions

## 4.2 Classification Tree

### 4.2.1 Assumptions

## 4.3 Summary of Findings

# 5 Conclusion

# 6 References

















